Viscous Transport in Eroding Porous Media

### 2019-10-15 docs/transport/erosion_transport.tex
modify figs in transport paper
giving it one more read through

Owing to the scales present in groundwater flow~\citep{bea1972}, we
model the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\citep{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the fluid
equations are linear and homogeneous, they are converted to a boundary
integral equation (BIE). Therefore, we resolve the non-negligible
interactions between nearly-touching arbitrarily shaped bodies, and the
erosion rate is computed accurately. We also compute the vorticity in
the fluid bulk since, on solid boundaries, vorticity reduces to shear
and thus provides a convenient way to simultaneously visualize local
erosion rates and changes in the surrounding flow
(figure~\ref{fig:Eroding50vort}).

@@

To compute stable simulations of erosion, we use methods of high-order
in both space and time. The time integration is unchanged from previous
work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be
triggered by changes in sign of the shear stress, and we use a stable
second-order Runge-Kutta method applied to the {\thL}
coordinates~\citep{hou-low-she1994} of the eroding grains.  In this
work, we introduce a new quadrature method to resolve dense suspensions.
The accuracy of the trapezoid rule, which was used in previous work, is
adequate for bodies that are sufficiently separated~\citep{tre-wei2014},
but not for grains in close contact.  One of the earliest quadrature
methods for nearly-singular integrands was developed
by~\citet{bak-she1986}, and in recent years, many other schemes have
followed~\citep{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\citep{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate the shear stress and
the fluid vorticity.

To characterize transport through the resulting configurations, particle
trajectories must be computed. Depending on the application, microscale
transport can be modelled as pure advection~\citep{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\citep{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\citep{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term, we solve~\eqref{eqn:tracers} with a fourth-order explicit
Runge-Kutta time stepping method. The Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain.

Once the streamlines are computed, we characterize transport by
analyzing three different metrics: the tortuosity, the anomalous
dispersion, and the pore size distribution.  The local tortuosity of a
streamline that connects the inlet to the outlet is defined as the
streamline's length normalized by the linear inlet-to-outlet distance.
In porous media, the local tortuosity can be greater than
1.5~\citep{kop-kat-tim1996, mat-kha-koz2008} or even
2~\citep{dud-koz-mat2011}, depending on several factors such as the
porosity. The tortuosity of a geometry is defined by averaging the local
tortuosity over all streamlines initialized at the inlet, and the
geometry's tortuosity characterizes average particle
motions~\citep{hak-com-den2019}.  To characterize spreading, the fluid
dispersion is computed by calculating the variance of the streamline
lengths. In porous media, this spreading is often
super-dispersive~\citep{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. Since anomalous dispersion results from
streamlines spending time in both the high and low velocity
regimes~\citep{ber-sch2001}, it is crucial to accurately resolve
streamlines near grain boundaries, as achieved in this work.  Finally,
we construct the pore-size distribution throughout the erosion process.
These distributions are required to quantify velocity
distributions~\citep{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\citep{sie-ili-pri-riv-gua2019},
connectivity~\citep{knu-car2005, wes-blo-gra2001}, and to develop
network models~\citep{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  We only
place eroding bodies in $[-1,1] \times [-1,1]$ to create a buffer region
that allows the flow profile imposed at the inlet to transition to the more
complex flow intervening between the bodies. On each eroding body, $\nn$
is the normal vector pointing into the body, and $\ss$ is the unit
tangent vector pointing in the counterclockwise direction.  The normal
vector on $\Gamma$ is the outward unit normal.  Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$. We simulate erosion by alternating
between solving the fluid equations and advancing the eroding grains.
The strength of $\UU$ is adjusted at each time step to achieve a
constant pressure drop across the channel, motivated by the geological
situation of a porous medium connecting two regions of  fixed hydraulic
heads.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we require accurate solutions of the incompressible Stokes
equations in complex two-dimensional geometries, we reformulate the
equations as a BIE.  This has the advantage that only the
one-dimensional boundary of the domain must be discretized, and, with
appropriate quadrature formulas and fast summation methods, the result
is a high-fidelity numerical simulation with near-optimal computational
complexity.

@@

Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, the
corresponding deformation tensor, pressure, and vorticity at $\xx \in
\Omega$ is written in terms of a layer potential~\cite{qua-moo2018}. To
compute the deformation tensor for $\xx \in \gamma$, we include the jump
term
\begin{align}
  \frac{1}{2} \left(\pderiv{\eeta}{\ss} \cdot \ss \right) \left[
    \begin{array}{cc}
      s_x^2 - s_y^2 & 2s_x s_y \\ 2s_x s_y & s_y^2 - s_x^2
    \end{array}
  \right].
  \label{eqn:deformationJump}
\end{align}
Finally, the deformation tensor, pressure, and vorticity due to the
Stokeslets and rotlets are readily available~\citep{poz1992}. Having
computed the deformation tensor on $\gamma$, the shear stress is
computed using equation~\eqref{eqn:shearStress}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cauchy Integral Representation of the Double-Layer
Potential}
\label{sec:DLPcomplex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The velocity double-layer potential~\eqref{eqn:velocityDLP}, and its
corresponding deformation tensor, pressure, and vorticity are all
written as layer potentials in $\RR^2$.  However, the quadrature method
we introduce in section~\ref{sec:method} requires a complex-valued
representation of the layer potentials for the velocity, deformation
tensor, and vorticity. The first step to form a complex representation
is to write the Laplace double-layer potential as the complex integral
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{2\upi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2}\eeta(\yy)\, ds_\yy = \Real (v(x)),
\end{align}
where
\begin{align}
  v(x) = \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{\eta(y)}{x - y} \, dy, \quad x \in \Omega.
  \label{eqn:laplaceComplex}
\end{align}
Here $x = x_1 + i x_2,y = y_1 + i y_2 \in \CC$ are the complex
counterparts of $\xx = (x_1,x_2),\yy = (y_1,y_2) \in \RR^2$, and $\eta =
\eta_1 + i \eta_2$ is the complex counterpart of $\eeta =
(\eta_1,\eta_2)$. Therefore, depending on the formulation of the layer
potential, $\Omega$ is being interpreted as a subset of $\RR^2$ or
$\CC$. Equation~\eqref{eqn:laplaceComplex} is converted to a Cauchy
integral by first finding the boundary data of $v$.  Assuming that
$\Omega$ is a simply-connected interior domain, the boundary data of $v$
satisfies the Sokhotski-Plemelj jump relation
\begin{align}
  \label{eqn:SPrelation}
  v(x) = - \frac{1}{2} \eta(x) + \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{\eta(y)}{x-y}\, dy, \quad x \in \Omega.
\end{align}
For exterior domains, the jump term changes from $-1/2$ to $1/2$, and
for multiply-connected domains, such as a porous media, $\bd\Omega$ is
decomposed into its different connected components and the appropriate
jump relation is applied.  Having computed the boundary data of the
holomorphic function $v$, by the Cauchy integral theorem we have
\begin{subequations}
  \label{eqn:cauchy}
  \begin{alignat}{3}
  v(x) &= \frac{1}{2\upi i}\int_{\bd\Omega} 
    \frac{v(y)}{y-x} \,dy, \\
  v'(x) &= \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{v(y)}{(y-x)^2} \, dy, \\
  v''(x) &= \frac{1}{\upi i} \int_{\bd\Omega}
    \frac{v(y)}{(y-x)^3} \, dy,
  \end{alignat}
\end{subequations}
for $x \in \Omega$.  Since $v(x)$ depends on the complex-valued density
function $\eta$, we use the notation $v[\eta](x)$ for the holomorphic
function defined in equation~\eqref{eqn:laplaceComplex}, and its first
two derivatives are written as $v'[\eta](x)$ and $v''[\eta](x)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Transport, Tracers, and Tortuosity}
\label{sec:transport}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Erosion in porous media leads to particular phenomena such as
channelization~\citep{berhanu2012shape}, and we are interested in
characterizing transport in such geometries.  In our previous
work~\citep{qua-moo2018}, we examined the effect of erosion on the area
fraction, flow rate, and the total drag.  However, to characterize
macroscopic signatures of the transport, other quantities must be
examined.  Here, we compute the anomalous dispersion rate, the
tortuosity, and the distribution of the pore sizes.  The first two
metrics can be defined in terms of streamlines governed by the
autonomous advection equation~\eqref{eqn:tracers}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Anomalous Dispersion}
\label{sec:dispersion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The spreading of fluid in a porous media is often characterized in terms
of anomalous dispersion~\citep{kla-rad-sok2008, den-cor-sch-ber2004}.
The anomalous dispersion rate depends on the porosity and
permeability~\citep{koc-bra1988}, but is also affected by the
distribution and shape of the grains. We calculate the anomalous
dispersion rates by analyzing the streamlines governed by
equation~\eqref{eqn:tracers} in eroded geometries. Given a set of $N_p$
trajectories, we define $\lambda_j(t)$ to be the arclength of the
trajectory
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss'_j(\tilde{t})\|\, d\tilde{t}, 
    \quad j=1,\ldots,N_p.
\end{align}
Then, the first and second ensemble moments are
\begin{align}
  \label{eqn:moments}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2,
\end{align}
and $\sigma_\lambda$ characterizes particle dispersion.  At early times,
the particles have not explored much of the geometry, and we expect a
ballistic motion $\sigma_\lambda \sim t$.  However, as the particles
pass the grains, their trajectories are altered, and we expect that
$\sigma_\lambda \sim t^\alpha$, with $\alpha \in (0.5,1)$, indicating
that the flow is super-dispersive.

To establish an anomalous dispersion rate, the trajectories must pass
several grains so that an asymptotic regime is reached. The geometries
that we consider are too short to observe asymptotic dispersion, so we
use a reinsertion method to form longer trajectories. Similar to the
work of others~\citep{dea-qua-bir-jua2018, puy-gou-den2019}, once a
particle reaches the outlet of the porous region, it is reinserted at
the inlet.  To minimize errors caused by reinsertion, the particle is
initialized at the discretization point that has the closest velocity to
the particle's velocity at the outlet.  After a single trajectory is
formed, it has undergone a collection of reinsertions.  Then, as a
post-processing step, the trajectory is made continuous by attaching the
tail of each segment to the origin of the next segment.

We consider trajectories in an open channel so that they can be compared
to trajectories in eroded geometries with large permeability.  The
trajectories are $(s_1(t),s_2(t)) = ((1-s_2(0)^2)t + s_1(0), s_2(0))$.
%\begin{align}
%  s_1(t) = (1-s_2(0)^2)t + s_1(0), \quad
%  s_2(t) = s_2(0).
%\end{align}
Averaging over $s_2(0) \in [-1,1]$, the first and second ensemble
moments are
\begin{align}
  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
\end{align}
In section~\ref{sec:results}, we compare the second moment in an open
channel with the trajectories in an eroded geometry.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial Discretization}
\label{sec:spatialDiscretization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we use a BIE formulation, we only need to discretize the
one-dimensional boundary of the domain.  Therefore, we easily resolve
complex geometries, and differentiation is performed with spectral
accuracy in Fourier space.  We discretize each eroding grain $\gamma_k$
with $N_\iin$ points and discretize the outer wall $\Gamma$ with
$N_\out$ points.  The $j^{th}$ discretization point on $\Gamma$ and
$\gamma_k$ are denoted by $\yy_j^0$ and $\yy^k_j$, respectively.  The
discretization points are initially distributed evenly in arclength, and
this equispacing is maintained throughout the entire simulation by using
the {\thL} formulation.  In addition, we apply
regularization~\citep{qua-moo2018} to slightly smooth the corners that
inevitably develop during erosion.

>>>>

Owing to the scales present in groundwater flow~\citep{bea1972}, we
model the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\citep{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the fluid
equations are linear and homogeneous, they are converted to a boundary
integral equation (BIE), and this allows us to naturally resolve the
non-negligible interactions between bodies.  We also compute the
vorticity in the fluid bulk since, on solid boundaries, vorticity
reduces to shear and thus provides a convenient way to simultaneously
visualize local erosion rates and changes in the surrounding flow
(figure~\ref{fig:Eroding50vort}).

@@

To compute stable simulations of erosion, we use methods of high-order
in both space and time. The time integration is unchanged from previous
work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be
triggered by changes in sign of the shear stress, and we use a stable
second-order Runge-Kutta method applied to the {\thL}
coordinates~\citep{hou-low-she1994} of the eroding grains.  In this
work, we introduce a new quadrature method to resolve dense suspensions.
The accuracy of the trapezoid rule, which was used in previous work, is
adequate for bodies that are sufficiently separated~\citep{tre-wei2014},
but not for grains in close contact.  One of the earliest quadrature
methods for nearly-singular integrands was developed
by~\citet{bak-she1986}, and in recent years, many other schemes have
followed~\citep{kli-tor2018, hel-oja2008a, bea-yin-wil2016, bea-lai2001,
klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\citep{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate the shear stress and
the fluid vorticity.

To characterize transport through the resulting configurations, particle
trajectories must be computed. Depending on the application, microscale
transport can be modelled as pure advection~\citep{dea-qua-bir-jua2018,
cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\citep{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\citep{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term, we solve~\eqref{eqn:tracers} with a fourth-order explicit
Runge-Kutta time stepping method. The Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain.

Once the streamlines are computed, we characterize transport by
analyzing three different metrics: the tortuosity, the anomalous
dispersion, and the pore size distribution.  The local tortuosity of a
streamline that connects the inlet to the outlet is defined as the
streamline's length normalized by the linear inlet-to-outlet distance.
In porous media, the local tortuosity can be greater than
1.5~\citep{kop-kat-tim1996, mat-kha-koz2008} or even
2~\citep{dud-koz-mat2011}, depending on several factors such as the
porosity. The tortuosity of a geometry is defined by averaging the local
tortuosity over all streamlines initialized at the inlet, and the
geometry's tortuosity characterizes average particle
motions~\citep{hak-com-den2019}.  To characterize spreading, the fluid
dispersion is defined as the variance of the streamline lengths. In
porous media, this spreading is often
super-dispersive~\citep{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. Since anomalous dispersion results from
streamlines spending time in both the high and low velocity
regimes~\citep{ber-sch2001}, it is crucial to accurately resolve
streamlines near grain boundaries, as achieved in this work.  Finally,
we construct the pore-size distribution throughout the erosion process.
These distributions are required to quantify velocity
distributions~\citep{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\citep{sie-ili-pri-riv-gua2019},
connectivity~\citep{knu-car2005, wes-blo-gra2001}, and to develop
network models~\citep{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  All eroding
bodies are placed in $[-1,1] \times [-1,1]$ to create a buffer region
that allows the flow profile imposed at the inlet to transition to the
more complex flow intervening between the bodies. Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, and $\Vn$ is the normal
velocity of $\gamma$. The shear stress on $\gamma$ is
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
where, $\nn$ is the normal vector pointing into the body, and $\ss$ is
the unit tangent vector pointing in the counterclockwise direction. We
simulate erosion by alternating between solving the fluid equations and
advancing the eroding grains.  The strength of $\UU$ is adjusted at each
time step to achieve a constant pressure drop across the channel,
motivated by the geological situation of a porous medium connecting two
regions of  fixed hydraulic heads.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To accurately solve the governing equations~\eqref{eqn:erosionModel} in
complex two-dimensional geometries, we reformulate the equations as a
BIE.  This has the advantage that only the one-dimensional boundary of
the domain must be discretized, and, with appropriate quadrature
formulas and fast summation methods, the result is a high-fidelity
numerical simulation with near-optimal computational complexity.

@@

Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, the
corresponding deformation tensor, pressure, and vorticity at $\xx \in
\Omega$ is written in terms of a layer potential~\citep{qua-moo2018}. To
compute the deformation tensor for $\xx \in \gamma$, we include the jump
term
\begin{align}
  \frac{1}{2} \left(\pderiv{\eeta}{\ss} \cdot \ss \right) \left[
    \begin{array}{cc}
      s_x^2 - s_y^2 & 2s_x s_y \\ 2s_x s_y & s_y^2 - s_x^2
    \end{array}
  \right].
  \label{eqn:deformationJump}
\end{align}
Finally, the deformation tensor, pressure, and vorticity due to the
Stokeslets and rotlets are readily available~\citep{poz1992}. Having
computed the deformation tensor on $\gamma$, the shear stress is
computed using equation~\eqref{eqn:shearStress}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cauchy Integral Representation of the Double-Layer
Potential}
\label{sec:DLPcomplex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The velocity double-layer potential~\eqref{eqn:velocityDLP}, and its
corresponding deformation tensor, pressure, and vorticity are all
written as layer potentials in $\RR^2$.  However, the quadrature method
we introduce in section~\ref{sec:method} requires complex-valued
representations. The first step to form a complex representation is to
write the Laplace double-layer potential as the complex integral
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{2\upi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2}\eeta(\yy)\, ds_\yy = \Real (v(x)),
\end{align}
where
\begin{align}
  v(x) = \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{\eta(y)}{x - y} \, dy, \quad x \in \Omega.
  \label{eqn:laplaceComplex}
\end{align}
Here $x = x_1 + i x_2,y = y_1 + i y_2 \in \CC$ are the complex
counterparts of $\xx = (x_1,x_2),\yy = (y_1,y_2) \in \RR^2$, and $\eta =
\eta_1 + i \eta_2$ is the complex counterpart of $\eeta =
(\eta_1,\eta_2)$. Therefore, depending on the formulation of the layer
potential, $\Omega$ is interpreted as a subset of $\RR^2$ or $\CC$.
Equation~\eqref{eqn:laplaceComplex} is converted to a Cauchy integral by
first finding the boundary data of $v$. If $\Omega$ is a
simply-connected interior domain, then the boundary data of $v$
satisfies the Sokhotski-Plemelj jump relation
\begin{align}
  \label{eqn:SPrelation}
  v(x) = - \frac{1}{2} \eta(x) + \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{\eta(y)}{x-y}\, dy, \quad x \in \bd\Omega.
\end{align}
For exterior domains, the jump term changes from $-1/2$ to $1/2$, and
for multiply-connected domains, such as a porous media, $\bd\Omega$ is
decomposed into its different connected components and the appropriate
jump relation is applied.  Having computed the boundary data of the
holomorphic function $v$, by the Cauchy integral theorem we have
\begin{subequations}
  \label{eqn:cauchy}
  \begin{alignat}{3}
  v(x) &= \frac{1}{2\upi i}\int_{\bd\Omega} 
    \frac{v(y)}{y-x} \,dy, \\
  v'(x) &= \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{v(y)}{(y-x)^2} \, dy, \\
  v''(x) &= \frac{1}{\upi i} \int_{\bd\Omega}
    \frac{v(y)}{(y-x)^3} \, dy,
  \end{alignat}
\end{subequations}
for $x \in \Omega$.  Since $v(x)$ depends on the complex-valued density
function $\eta$, we use the notation $v[\eta](x)$ for the holomorphic
function defined in equation~\eqref{eqn:laplaceComplex}, and its first
two derivatives are written as $v'[\eta](x)$ and $v''[\eta](x)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Transport, Tracers, and Tortuosity}
\label{sec:transport}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Erosion in porous media leads to phenomena such as
channelization~\citep{berhanu2012shape}, and we are interested in
characterizing transport in such geometries.  In our previous
work~\citep{qua-moo2018}, we examined the effect of erosion on the area
fraction, flow rate, and the total drag.  However, to characterize
macroscopic signatures of the transport, other quantities must be
examined.  Here, we compute the anomalous dispersion rate, the
tortuosity, and the distribution of the pore sizes.  The first two
metrics are defined in terms of streamlines governed by the autonomous
advection equation~\eqref{eqn:tracers}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Anomalous Dispersion}
\label{sec:dispersion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The spreading of fluid in a porous media is often characterized in terms
of anomalous dispersion~\citep{kla-rad-sok2008, den-cor-sch-ber2004}.
The anomalous dispersion rate depends on the porosity and
permeability~\citep{koc-bra1988}, but is also affected by the
distribution and shape of the grains. We calculate the anomalous
dispersion rates by analyzing the streamlines governed by
equation~\eqref{eqn:tracers} in eroded geometries. Given a set of $N_p$
trajectories, we define $\lambda_j(t)$ to be the arclength of the
trajectory
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss'_j(\tilde{t})\|\, d\tilde{t}, 
    \quad j=1,\ldots,N_p.
\end{align}
Then, the first and second ensemble moments are
\begin{align}
  \label{eqn:moments}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2,
\end{align}
and $\sigma_\lambda$ characterizes the dispersion.  At early times, the
particles have not explored much of the geometry, and we expect a
ballistic motion $\sigma_\lambda \sim t$.  However, as the particles
pass the grains, their trajectories are altered, and we expect that
$\sigma_\lambda \sim t^\alpha$, with $\alpha \in (0.5,1)$, indicating
that the flow is super-dispersive.

To establish an asymptotic anomalous dispersion rate, the trajectories
must pass several grains. The geometries that we consider are too short
to observe asymptotic dispersion, so we use a reinsertion method to form
longer trajectories. Similar to the work of
others~\citep{dea-qua-bir-jua2018, puy-gou-den2019}, once a particle
reaches the outlet of the porous region, it is reinserted at the inlet.
To minimize errors caused by reinsertion, the particle is initialized at
the discretization point that has the closest velocity to the particle's
velocity at the outlet.  After a single trajectory is formed, it has
undergone a collection of reinsertions.  Then, as a post-processing
step, the trajectory is made continuous by attaching the tail of each
segment to the origin of the next segment.

%We consider trajectories in an open channel so that they can be compared
%to trajectories in eroded geometries with large permeability.  The
%trajectories are $(s_1(t),s_2(t)) = ((1-s_2(0)^2)t + s_1(0), s_2(0))$.
%%\begin{align}
%%  s_1(t) = (1-s_2(0)^2)t + s_1(0), \quad
%%  s_2(t) = s_2(0).
%%\end{align}
%Averaging over $s_2(0) \in [-1,1]$, the first and second ensemble
%moments are
%\begin{align}
%  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
%    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
%\end{align}
%In section~\ref{sec:results}, we compare the second moment in an open
%channel with the trajectories in an eroded geometry.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial Discretization}
\label{sec:spatialDiscretization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we use a BIE formulation, we only need to discretize the
one-dimensional boundary of the domain.  We discretize each eroding
grain $\gamma_k$ with $N_\iin$ points and discretize the outer wall
$\Gamma$ with $N_\out$ points.  The $j^{th}$ discretization point on
$\Gamma$ and $\gamma_k$ are denoted by $\yy_j^0$ and $\yy^k_j$,
respectively.  The discretization points are initially distributed
evenly in arclength, and this equispacing is maintained throughout the
entire simulation by using the {\thL} formulation.  In addition, we
apply regularization~\citep{qua-moo2018} to slightly smooth the corners
that inevitably develop during erosion.
